Fractional Types

### 2019-11-16 fracGC/p.tex
paper

A conventional and efficient way to ensure that every allocation is
paired with a matching de-allocation is to impose a stack
discipline. An illustrative demonstration of this idea is the
\verb|with_ancilla| operator from
Quipper~\cite{Green:2013:QSQ:2491956.2462177}:
\begin{verbatim}
with_ancilla :: (Qubit -> Circ a) -> Circ a
\end{verbatim}
The operator takes a block of gates parameterized by an ancilla,
allocates a new ancilla initialized to $\ket{0}$, and runs the given
block of gates. At the end of its execution, the block is expected to
return the ancilla to the state $\ket{0}$ (this is the second
subproblem addressed later) at which point it is de-allocated. In this
scoped-by-construction approach, allocation and de-allocation of
ancillae requires nothing beyond conventional parameter-passing
techniques in which the parameter is allocated before entry to a
function and de-allocated at exit.

This scoped model is therefore quite a pragmatic choice. It is however
limited and to understand its limitations more vividly, we propose the
following analogy: allocating an ancilla by creating a new wire in the
circuit is like borrowing some money; the computation has access to a
new resource temporarily. De-allocating the ancilla is like returning
the borrowed money; the computation no longer has access to the
additional resource. It would be unreasonably restrictive to insist
that if the person (function) borrowing the money must be the same
person (function) who returns it, i.e., preventing the following
computation:
\begin{verbatim}
  |-----\  /-----|
         \/
        / \
  |----/   \-----|
\end{verbatim}
Indeed, as far as reversible or quantum computation is concerned, the
circuit above is perfectly sensible since it conserves information
(all money borrowed is repaid and the net balance is 0). In fact one
can go further, if one person borrows \$10, why couldn't three
different persons collectively collaborate to pay back the debt with
one person paying \$5, another \$2, and a third \$3. This extra
generality is not a gratuitous concern: since scope is fundamentally a
static property of programs, it does not allow the flexibility of heap
allocation in which the lifetime of resources is dynamically
determined. To summarize, since the reversible-quantum computational
framework guarantees that information is preserved, it also permits
fascinating mechanisms in which computations can be sliced and
diced, decomposed and recomposed, run forwards and backwards, in
arbitrary ways as long as the net balance is 0.

>>>>

A conventional and efficient way to ensure that every allocation is
paired with a matching de-allocation is to impose a stack
discipline. An illustrative demonstration of this idea is the
\verb|with_ancilla| operator from
Quipper~\cite{Green:2013:QSQ:2491956.2462177}:
\begin{verbatim}
with_ancilla :: (Qubit -> Circ a) -> Circ a
\end{verbatim}
The operator takes a block of gates parameterized by an ancilla,
allocates a new ancilla of type \verb|Qubit| initialized to $\ket{0}$,
and runs the given block of gates. At the end of its execution, the
block is expected to return the ancilla to the state $\ket{0}$ (this
is the second subproblem addressed later) at which point it is
de-allocated. In this scoped-by-construction approach, allocation and
de-allocation of ancillae requires nothing beyond conventional
parameter-passing techniques in which the parameter is allocated
before entry to a function and de-allocated at exit.

This scoped model is therefore quite a pragmatic choice. It is however
limited, and to understand its limitations more vividly, we propose
the following analogy: allocating an ancilla by creating a new wire in
the circuit is like borrowing some money from a global external
entity; the computation has access to a new resource
temporarily. De-allocating the ancilla is like returning the borrowed
money to the global entity; the computation no longer has access to
the additional resource. It would however be unreasonably restrictive
to insist that if the person (function) borrowing the money must be
the same person (function) returning it.
%% , i.e., preventing the following computation:
%% \begin{verbatim}
%%   |-----\  /-----|
%%          \/
%%         / \
%%   |----/   \-----|
%% \end{verbatim}
Indeed, as far as reversible or quantum computation is concerned, the
only important invariant is that information is conserved, i.e., that
money is conserved. The identities of bits or qubits are not
observable as they are all interchangeable in the same way that
particular bills with different serial numbers are interchangeable in
financial transactions. Thus the only invariant is that the net flow
of money between the computation and the global entity is zero. This
observation allows us to go even further than just switching the
identities of borrowers. It is even possible for one person to borrow
\$10, and have three different persons collectively collaborate to
pay back the debt with one person paying \$5, another \$2, and a third
\$3.

Computationally, this extra generality is not a gratuitous concern:
since scope is fundamentally a static property of programs, it does
not allow the flexibility of heap allocation in which the lifetime of
resources is dynamically determined. To summarize, since the
reversible-quantum computational framework guarantees that information
is preserved, it also permits fascinating mechanisms in which
computations can be sliced and diced, decomposed and recomposed, run
forwards and backwards, in arbitrary ways as long as the net balance
is 0.
